An exposition on the propriety of restricted Boltzmann machines

### 2016-09-12 writing/draft.Rmd
through toy model, blue edits

---
title: "An exposition on the propriety of restricted Boltzmann machines"
author: "Andee Kaplan, Daniel Nordman, Stephen Vardeman"
date: "August, 2016"
linestretch: 2 
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, with two layers, one hidden and one visible, with no conditional dependency within a layer. In recent years, RBMs have risen to prominence due to their connection to deep learning. By treating a hidden layer of one RBM as the visible layer in a second RBM a deep architecture can be created. RBMs are thought to have the ability to encode very complex and rich structures in data, making them attractive for supervised learning. However, the generative behavior of RBMs has largely been unexplored. In this paper, we discuss the relationship between parameter RBM specification and the prevalence of degenerate models as well as the difficulties that arise in fitting such highly flexible models in the binary case.
---

@@

Where the denominator $\gamma(\boldsymbol \theta)$ is the normalizing function that ensures the distribution sums to one, $\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)$. It is standard to refer to $Q(\boldsymbol x) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$ as the neg-potential function in this model. The RBM model is parameterized by $\boldsymbol \theta$, which contains two types of parameters, main effects and interaction effects. The main effects parameters ($\theta_{v_i}, \theta_{h_j}$) weight the values of the visible and hidden nodes and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections, or dependencies, between hidden and visible layers.

@@

The RBM class of models can be thought to exhibit an analogous occurence of *near-degeneracy* when random variables in the neg-potential function $Q(\cdot)$, $t(\boldsymbol x)$, have a mean vector $\mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{X}$, where $t(\boldsymbol x) = \{v_1, \dots, v_V, h_1, \dots, h_H, v_1 h_1, \dots, v_V h_H \} \in \mathcal{X}$ and the mean parameterization on the model parameters, $\mu(\boldsymbol \theta)$ is
\begin{align*}
\mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta} t(\boldsymbol X)  \\
&= \sum\limits_{x \in \mathcal{X}} \left\{ t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} \\
&= \sum\limits_{x \in \mathcal{X}} \left\{ t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}\right\} \\
\end{align*}

@@

@schweinberger2011instability introduces a related concept, *instability*. Instability is characterized by excessive sensitivity in the model, where a small change in $\boldsymbol x$ leads to a large change in the probability mass function. Consider discrete exponential families of distributions with probability mass functions of the form 
\begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}
\end{align*}
where, $\eta(\boldsymbol \theta)$ is a vector of natural parameters, and $t: \mathcal{X} \rightarrow \mathbb{R}^q$ is a vector of sufficient statistics based on $\boldsymbol x \in \mathcal{X}^N$ for $\mathcal{X}$ a discrete set of $M$ elements and $N$ degrees of freedom. In the random graph application, $\mathcal{X}^N$ may correspond to the number of relationships among $n$ nodes in a graph. In the instance of RBMs, $N$ refers to the number of hiddens and visible nodes included in the model, $N = H + V$. Again, $\gamma(\boldsymbol \theta)$ is the normalizing function that guarantees the distribution is valid, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}^N} \exp\left(\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right))$.

@schweinberger2011instability provides the following formal definition for instability in a discrete exponential family model.

\begin{defn}
A discrete exponential family distribution is \emph{unstable} if for any $C > 0$, however large, there exists $N_C > 0$ such that 
\begin{align}
\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right)] - \min\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right)] > CN \text{ for all } N > N_C \label{eq:unstable}
\end{align}

Conversely, a discrete exponential family distribution is \emph{stable} if there exists constants $C > 0$ and $N_C > 0$ such that 
\begin{align}
\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right)] - \min\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right)] \le CN \text{ for all } N > N_C. \label{eq:stable}
\end{align}
\end{defn}

For RBMs, an analagous concept could concern the joint distribution of both hidden and visible nodes, where $N$ is the number of nodes in the model, $H + V$. However, instability in terms of only the visible nodes is a more useful measure of model impropriety. To see this, consider modeling images using an RBM. One would be interested in the maximum change in the likelihood when a single pixel is changed in the visibles as a proxy for instability in the model. So we concern ourselves with a notion of instability that deals with the maximum likelihood ratio of two vectors of visibless that differ at only one coordinate.

@@

Instability in a discrete exponential model is in itself a measure of model impropriety, however there is also a relationship between instability and near-degeneracy in a discrete exponental model.

\begin{lemma}[Schweinberger (2011)]
If a discrete exponential family distribution is unstable, then it is degenerate in the sense that for any $0 < \epsilon < 1$, $P_\theta(\boldsymbol X \in \mathcal{M}_{\epsilon, N}) \rightarrow 1$ as $N \rightarrow \infty$ where $\mathcal{M}_{\epsilon, N} = \{boldsymbol x \in \mathcal{X}^N:  \eta^T(\boldsymbol \theta) t(\boldsymbol x)\right) > (1-\epsilon)\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right)]$ is the set of points in the sample space that yield arbitrarily high values of the unnormalized pmf.
\end{lemma}

Thus, if a discrete exponential family model is unstable, then it is also near-degenerate. Discrete RBMs (the focus of this paper) are curved exponential family models for particular choice of sufficient statistics ($\{h_j, v_i, v_i h_j: i = 1, \dots, V, j= 1, \dots, H\}$), meaning instability in the RBM implies near-degeneracy as well.

@@

## Tiny example
To illustrate the ideas of graph near-degeneracy, instability, and uninterpretability in a RBM, we consider first the smallest possible (toy) example that consists of one visible node $v_1$ and one hidden node $h_1$, both binary. A schematic of this model can be found in figure \ref{fig:toymodel}.

@@

For this small model, we are able to investigate the symptoms of the three measure of model impropriety, near-degeneracy, instability uninterpretability. First, we are able to illustrate the shape of boundary of the convex hull of $\{t_k(\boldsymbol x): \boldsymbol x \in \mathcal{X}, k = 1, \dots, K\}$ and explore the behavior of the mean parameterization of the parameters because we are in three dimensions. In figure \ref{fig:toyhull}, the convex hull of our statistic space in three dimensions for the toy RBM with one visible and one hidden node from multiple perspectives enclosed by an unrestricted hull of $\{0,1\}^3$ space is shown. In this small model, $\{t(\boldsymbol x)\}$ is defined as $\{t(\boldsymbol x)\} = \{v_1, h_1, v_1 h_1\}$ by definition and form of the neg-potential function $Q(\boldsymbol x)$. Note that the convex hull of $\{t(\boldsymbol x)\}$ does not fill the unrestricted hull because of the relationship between the elements of $\{t(\boldsymbol x)\}$.

@@

    ```{r, results='asis', echo=FALSE, include=TRUE, message=FALSE}
    cat(sapply(1:nrow(m1_bin$c_hull$hull), function(x) paste0("\\draw", paste(paste0("(", apply(m1_bin$possible_t[cbind(m1_bin$c_hull$hull, m1_bin$c_hull$hull[,1])[x,],], 1, paste, collapse=","), ")"), collapse=" -- "), "; \n")))
    ```
    \end{tikzpicture}
  }
  \end{minipage}
  \caption{The convex hull of our statistic space in three dimensions for the toy RBM with one visible and one hidden node from multiple perspectives enclosed by an unrestricted hull of $\{0,1\}^3$ space. Note that the convex hull of $\{t(\boldsymbol x)\}$ does not fill the unrestricted hull because of the relationship between the elements of $\{t(\boldsymbol x)\}$.}
\label{fig:toyhull}
\end{figure}

Additionally, we can compute the mean parametrization of the model parameters as 
\begin{align*}
\mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta}\left[ t(X) \right] \\
&= \sum\limits_{x \in \mathcal{X}} \left\{ t(x) \frac{\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)} \right\} \\
&= \sum\limits_{x \in \mathcal{X}} \left\{ t(x) \frac{\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)}{\gamma(\boldsymbol \theta)} \right\} \\
&= \left[
\begin{matrix}
\frac{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\gamma(\boldsymbol \theta)} \\
\frac{\exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\gamma(\boldsymbol \theta)} \\
\frac{\exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\gamma(\boldsymbol \theta)} \\
\end{matrix} \right]
\end{align*}
These three functions can be visualized over multiple values of $\boldsymbol \theta$, as seen in figure \ref{fig:degen_toy}. By examining these functions, we can see that as values of $\boldsymbol \theta$ grow larger in magnitude, at least one mean-parameterization functions is approaching the values 0 or 1, indicating a potential to be near to the boundary of the convex hull of $\{t_k(\boldsymbol x): \boldsymbol x \in \mathcal{X}, k = 1, \dots, K\}$. Thus, for a very small example we can see the relationship between values of $\boldsymbol \theta$ and model degeneracy.

```{r near_dependency, fig.cap="\\label{fig:degen_toy}The three mean-parameterization functions of the model parameters for a RBM with one visible and one hidden node.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE}
possibles <- stats(1, 1, type = "binary")

expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
            theta_v1 = seq(-5, 5, length.out = 21),
            theta_11 = seq(-5, 5, length.out = 5)) %>%
  data.frame() %>%
  group_by(theta_v1, theta_h1, theta_11) %>%
  do(data.frame(gamma = sum(exp(possibles %*% t(data.matrix(.)))),
                mu_v1 = possibles[,"v1"] %*% exp(possibles %*% t(data.matrix(.))),
                mu_h1 = possibles[,"h1"] %*% exp(possibles %*% t(data.matrix(.))),
                mu_11 = possibles[,"theta11"] %*% exp(possibles %*% t(data.matrix(.))))) %>%
  mutate_each(funs(./gamma), starts_with("mu")) %>%
  select(-gamma) %>%
  gather(mu_parameter, value, mu_v1, mu_h1, mu_11) %>%
  ungroup() %>%
  mutate(theta_11 = factor(theta_11, labels = paste0("theta[11]==", unique(theta_11))),
         mu_parameter = factor(mu_parameter, labels = paste0(gsub("_", "(theta[", unique(mu_parameter)), "])"))) %>%
  ggplot() +
  geom_tile(aes(theta_v1, theta_h1, fill = value))  +
  geom_contour(aes(theta_v1, theta_h1, z = value), colour = "black") +
  facet_grid(mu_parameter ~ theta_11, labeller = label_parsed) +
  xlab(expression(theta[v1])) +
  ylab(expression(theta[h1])) +
  scale_fill_gradient2("Mean-parameterization", low = "#af8dc3", high = "#7fbf7b", midpoint = .5, limits = c(0, 1)) +
  theme(aspect.ratio = 1)
```

Secondly, we can look at $R(\boldsymbol \theta)$ for this tiny model, which gives us a lower bound for the loglikelihood ratio of the smallest possible change in $\boldsymbol x$, indicating instability.

@@

Figure \ref{fig:instab} shows $R(\boldsymbol \theta)/V$ for various values of $\boldsymbol \theta$. We can see that this quantity is increasing for larger magnitudes of $\boldsymbol \theta$, especially for large values of $\theta_11$. This may indicate instability as $|\boldsymbol \theta|$ is large, agreeing with the indications about near-degeneracy from looking at $\mu(\boldsymbol \theta)$.

```{r instab, fig.cap="\\label{fig:instab}$R(\\boldsymbol \\theta)/V$ for various values of $\\boldsymbol \\theta$ for the tiny example model. This quantity is increasing for larger positive values of $\\boldsymbol \\theta$.", fig.height = 4, fig.align='center'}
expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
            theta_v1 = seq(-5, 5, length.out = 21),
            theta_11 = seq(-5, 5, length.out = 5)) %>%
  data.frame() %>%
  group_by(theta_v1, theta_h1, theta_11) %>%
  do(data.frame(Q = cbind(possibles, possibles %*% t(data.matrix(.))))) %>%
  group_by(theta_v1, theta_h1, theta_11, Q.v1) %>%
  do(data.frame(maxq = max(.$Q.V4))) %>%
  group_by(theta_v1, theta_h1, theta_11) %>%
  do(data.frame(q_max = max(.$maxq) - min(.$maxq) - log(2))) %>%
  ungroup() %>%
  mutate(theta_11 = factor(theta_11, labels = paste0("theta[11]==", unique(theta_11)))) %>%
  ggplot() +
  geom_tile(aes(theta_v1, theta_h1, fill = q_max)) +
  geom_contour(aes(theta_v1, theta_h1, z = q_max), colour = "black") +
  facet_grid(. ~ theta_11, labeller = label_parsed) +
  xlab(expression(theta[v1])) +
  ylab(expression(theta[h1])) +
  scale_fill_gradient(expression(frac(R(theta), V)), low = "yellow", high = "red") +
  theme(aspect.ratio = 1)
```

Finally, we can look at the difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \emptyset\right ]$ for the tiny toy model to understand the effect of uninterpretability on the RBM.

>>>>

---
title: "An exposition on the propriety of restricted Boltzmann machines"
author: "Andee Kaplan, Daniel Nordman, Stephen Vardeman"
date: "September, 2016"
linestretch: 2 
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, with two layers, one hidden and one visible, with no conditional dependency within a layer. In recent years, RBMs have risen to prominence due to their connection to deep learning. By treating a hidden layer of one RBM as the visible layer in a second RBM a deep architecture can be created. RBMs are thought to have the ability to encode very complex and rich structures in data, making them attractive for supervised learning. However, the generative behavior of RBMs has largely been unexplored. In this paper, we discuss the relationship between parameter RBM specification and the prevalence of degenerate models as well as the difficulties that arise in fitting such highly flexible models in the binary case.
---

@@

Where the denominator $\gamma(\boldsymbol \theta)$ is the normalizing function that ensures the distribution sums to one, $\gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)$. Let $\boldsymbol \tau = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\} = \{v_i, h_j, v_i h_j: i = 1, \dots, V, j = 1, \dots H\}$ be the vector of sufficient statistics in the model and $Q(\boldsymbol x) = \sum\limits_{i = 1}^H\sum\limits_{j=1}^V \theta_{ij} h_i v_j + \sum\limits_{i = 1}^H\theta_{hi} h_i + \sum\limits_{j = 1}^V\theta_{vj} v_j$, the neg-potential function. The RBM model is parameterized by $\boldsymbol \theta$, which contains two types of parameters, main effects and interaction effects. The main effects parameters ($\theta_{v_i}, \theta_{h_j}$) weight the values of the visible and hidden nodes and the interaction effects parameters ($\theta_{ij}$) weight the values of the connections, or dependencies, between hidden and visible layers.

@@

The RBM class of models can be thought to exhibit an analogous occurrence of *near-degeneracy* when random variables in the neg-potential function $Q(\cdot)$, $t(\boldsymbol x)$, have a mean vector $\mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{X}$, where $t(\boldsymbol x) = \{v_1, \dots, v_V, h_1, \dots, h_H, v_1 h_1, \dots, v_V h_H \} \in \mathcal{X}$ and the mean parameterization on the model parameters, $\mu(\boldsymbol \theta)$ is
\begin{align*}
\mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta} t(\boldsymbol X)  \\
&= \sum\limits_{x \in \mathcal{X}} \left\{ t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} \\
&= \sum\limits_{x \in \mathcal{X}} \left\{ t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}\right\} \\
\end{align*}

@@

@schweinberger2011instability introduces a related concept, *instability*. Instability is characterized by excessive sensitivity in the model, where a small change in $\boldsymbol x$ leads to a large change in the probability mass function. Consider discrete exponential families of distributions with probability mass functions of the form 
\begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\eta^T(\boldsymbol \theta) t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}
\end{align*}
@schweinberger2011instability provides the following formal definition for instability in a discrete exponential family model.

\begin{defn}
A discrete exponential family distribution is \emph{unstable} if for any $C > 0$, however large, there exists $N_C > 0$ such that 
\begin{align}
\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)] - \min\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)] > CN \text{ for all } N > N_C \label{eq:unstable}
\end{align}

Conversely, a discrete exponential family distribution is \emph{stable} if there exists constants $C > 0$ and $N_C > 0$ such that 
\begin{align}
\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)] - \min\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)] \le CN \text{ for all } N > N_C. \label{eq:stable}
\end{align}
\end{defn}

For RBMs, an analogous concept could concern the joint distribution of both hidden and visible nodes, where $N$ is the number of nodes in the model, $H + V$. However, instability in terms of only the visible nodes is a more useful measure of model impropriety. To see this, consider modeling images using an RBM. One would be interested in the maximum change in the likelihood when a single pixel is changed in the visibles as a proxy for instability in the model. So we concern ourselves with a notion of instability that deals with the maximum likelihood ratio of two vectors of visibles that differ at only one coordinate.

@@

Instability in a discrete exponential model is in itself a measure of model impropriety, however there is also a relationship between instability and near-degeneracy in a discrete exponential model.

\begin{lemma}[Schweinberger (2011)]
If a discrete exponential family distribution is unstable, then it is degenerate in the sense that for any $0 < \epsilon < 1$, $P_\theta(\boldsymbol X \in \mathcal{M}_{\epsilon, N}) \rightarrow 1$ as $N \rightarrow \infty$ where $\mathcal{M}_{\epsilon, N} = \{\boldsymbol x \in \mathcal{X}^N:  \eta^T(\boldsymbol \theta) t(\boldsymbol x) > (1-\epsilon)\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[\eta^T(\boldsymbol \theta) t(\boldsymbol x)]\}$ is the set of points in the sample space that yield arbitrarily high values of the unnormalized pmf.
\end{lemma}

Thus, if a discrete exponential family model is unstable, then it is also near-degenerate. Discrete RBMs (the focus of this paper) are curved exponential family models for particular choice of sufficient statistics ($\boldsymbol \tau = \{h_j, v_i, v_i h_j: i = 1, \dots, V, j= 1, \dots, H\}$), meaning instability in the RBM implies near-degeneracy as well.

@@

## Tiny example
To illustrate the ideas of model near-degeneracy, instability, and uninterpretability in a RBM, we consider first the smallest possible (toy) example that consists of one visible node $v_1$ and one hidden node $h_1$ that are both binary. A schematic of this model can be found in figure \ref{fig:toymodel}. Because it seems most common, we'll begin by employing $0/1$ encoding of binary variables (both $h_1$ and $v_1$ taking values in $\{0,1\}$). Eventually we'll note in section \ref{data-coding-to-mitigate-degeneracy} that $-1/1$ coding has substantial advantages.

@@

For this small model, we are able to investigate the symptoms of model impropriety (near-degeneracy, instability uninterpretability). First, because we are in three dimensions, we are able to explicitly illustrate the shape of boundary of the convex hull of $\boldsymbol tau$ and explore the behavior of the mean parameterization of the parameters. In figure \ref{fig:toyhull}, the convex hull of our "statistic space", $\boldsymbol \tau$, in three dimensions for this toy problem is shown from multiple perspectives enclosed by the unit cube, it convex hull of $\{0,1\}^3$ space. In this small model, note that the convex hull of $\boldsymbol tau$ does not fill the unrestricted hull because of the relationship between the elements of $\{(v_1, h_1, v_1 h_1 : (v_1, h_1) \in \{0, 1\}^2 \}$.

@@

    ```{r, results='asis', echo=FALSE, include=TRUE, message=FALSE}
    cat(sapply(1:nrow(m1_bin$c_hull$hull), function(x) paste0("\\draw", paste(paste0("(", apply(m1_bin$possible_t[cbind(m1_bin$c_hull$hull, m1_bin$c_hull$hull[,1])[x,],], 1, paste, collapse=","), ")"), collapse=" -- "), "; \n")))
    ```
    \end{tikzpicture}
  }
  \end{minipage}
  \caption{The convex hull of the "statistic space"" in three dimensions for the toy RBM with one visible and one hidden node. The convex hull of $\boldsymbol \tau = \{\boldsymbol t(\boldsymbol x) : \boldsymbol x \in \mathcal{X}\}$ does not fill the unit cube because of the relationship between the elements of $\boldsymbol \tau$.}
\label{fig:toyhull}
\end{figure}

Additionally, we can compute the mean vector for $\boldsymbol t(\boldsymbol x)$ as a function of the model parameters as 
\begin{align*}
\boldsymbol \mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta}\left[ \boldsymbol t(\boldsymbol X) \right] \\
&= \sum\limits_{\boldsymbol x \in \mathcal{X}} \left\{ \boldsymbol t(\boldsymbol x) \frac{\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)} \right\} \\
&= \sum\limits_{\boldsymbol x \in \mathcal{X}} \left\{ t(x) \frac{\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)}{\gamma(\boldsymbol \theta)} \right\} \\
&= \left[
\begin{matrix}
\frac{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\gamma(\boldsymbol \theta)} \\
\frac{\exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\gamma(\boldsymbol \theta)} \\
\frac{\exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\gamma(\boldsymbol \theta)} \\
\end{matrix} \right]
\end{align*}
The three parametric functions giving $\boldsymbol \mu(\boldsymbol \theta)$ can be visualized in figure \ref{fig:degen_toy}. By examining these, we see that as values of coordinates of $\boldsymbol \theta$ grow larger in magnitude, at least one mean function for the entries of $\boldsymbol t(\boldsymbol x)$ is approaching a value 0 or 1, indicating a potential for $\text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ to be near to the boundary of the convex hull of $\boldsymbol \tau$. Thus, for a very small example we can see the relationship between values of $\boldsymbol \theta$ and model degeneracy.

```{r near_dependency, fig.cap="\\label{fig:degen_toy}The three parametric mean functions of sufficient statistics for a RBM with one visible and one hidden node.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE}
possibles <- stats(1, 1, type = "binary")

expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
            theta_v1 = seq(-5, 5, length.out = 21),
            theta_11 = seq(-5, 5, length.out = 5)) %>%
  data.frame() %>%
  group_by(theta_v1, theta_h1, theta_11) %>%
  do(data.frame(gamma = sum(exp(possibles %*% t(data.matrix(.)))),
                mu_v1 = possibles[,"v1"] %*% exp(possibles %*% t(data.matrix(.))),
                mu_h1 = possibles[,"h1"] %*% exp(possibles %*% t(data.matrix(.))),
                mu_11 = possibles[,"theta11"] %*% exp(possibles %*% t(data.matrix(.))))) %>%
  mutate_each(funs(./gamma), starts_with("mu")) %>%
  select(-gamma) %>%
  gather(mu_parameter, value, mu_v1, mu_h1, mu_11) %>%
  ungroup() %>%
  mutate(theta_11 = factor(theta_11, labels = paste0("theta[11]==", unique(theta_11))),
         mu_parameter = ifelse(mu_parameter == "mu_v1", "paste(E[theta], v[1])", 
                               ifelse(mu_parameter == "mu_h1", "paste(E[theta], h[1])", "paste(E[theta], v[1], h[1])"))) %>%
  #factor(mu_parameter, labels = paste0(gsub("_", "(theta[", unique(mu_parameter)), "])"))) %>%
  ggplot() +
  geom_tile(aes(theta_v1, theta_h1, fill = value))  +
  geom_contour(aes(theta_v1, theta_h1, z = value), colour = "black") +
  facet_grid(mu_parameter ~ theta_11, labeller = label_parsed) +
  xlab(expression(theta[v1])) +
  ylab(expression(theta[h1])) +
  scale_fill_gradient2("Mean", low = "#af8dc3", high = "#7fbf7b", midpoint = .5, limits = c(0, 1)) +
  theme(aspect.ratio = 1)
```

Secondly, we can look at $R(\boldsymbol \theta)$ for this tiny model. Recall that gives us a lower bound for the loglikelihood ratio of the minimum possible change in $\boldsymbol x$, indicating instability.

@@

Figure \ref{fig:instab} shows $R(\boldsymbol \theta)/V$ for various values of $\boldsymbol \theta$. We can see that this quantity is increasing for larger magnitudes of $\boldsymbol \theta$, especially for large values of $\theta_{11}$. This suggests instability as $|\boldsymbol \theta|$ becomes large, agreeing with the concerns about near-degeneracy produced by consideration of $\boldsymbol \mu(\boldsymbol \theta)$.

```{r instab, fig.cap="\\label{fig:instab}$R(\\boldsymbol \\theta)/V$ for various values of $\\boldsymbol \\theta$ for the tiny example model. (Recall that here $V$ is the number of visible nodes and here is $1$. This quantity is increasing for larger magnitudes of $\\boldsymbol \\theta$.", fig.height = 4, fig.align='center'}
expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
            theta_v1 = seq(-5, 5, length.out = 21),
            theta_11 = seq(-5, 5, length.out = 5)) %>%
  data.frame() %>%
  group_by(theta_v1, theta_h1, theta_11) %>%
  do(data.frame(Q = cbind(possibles, possibles %*% t(data.matrix(.))))) %>%
  group_by(theta_v1, theta_h1, theta_11, Q.v1) %>%
  do(data.frame(maxq = max(.$Q.V4))) %>%
  group_by(theta_v1, theta_h1, theta_11) %>%
  do(data.frame(q_max = max(.$maxq) - min(.$maxq) - log(2))) %>%
  ungroup() %>%
  mutate(theta_11 = factor(theta_11, labels = paste0("theta[11]==", unique(theta_11)))) %>%
  ggplot() +
  geom_tile(aes(theta_v1, theta_h1, fill = q_max)) +
  geom_contour(aes(theta_v1, theta_h1, z = q_max), colour = "black") +
  facet_grid(. ~ theta_11, labeller = label_parsed) +
  xlab(expression(theta[v1])) +
  ylab(expression(theta[h1])) +
  scale_fill_gradient(expression(frac(R(theta), V)), low = "yellow", high = "red") +
  theme(aspect.ratio = 1)
```

Finally, we can look at the difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ for the tiny toy model to understand the effect of $\boldsymbol \theta$ on uninterpretability in the RBM.

\begin{align*}
\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] 
&= \left[
\begin{matrix}
\frac{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)} \\
\frac{\exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)} 
\end{matrix} \right] - \left[
\begin{matrix}
\frac{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{v_1} + \theta_{h_1}\right)}{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{v_1} + \theta_{h_1}\right)} \\
\frac{\exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)}{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)} 
\end{matrix} \right] \\
&= \left[
\begin{matrix}
\frac{\left|\exp\left(\theta_{11} + \theta_{v_1} + 2\theta_{h_1}\right) - \exp\left( \theta_{v_1} + 2\theta_{h_1}\right) \right| }{\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)\right)\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)\right)} \\
\frac{\left|\exp\left(\theta_{11} + 2\theta_{v_1} + \theta_{h_1}\right) - \exp\left( 2\theta_{v_1} + \theta_{h_1}\right) \right| }{\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)\right)\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)\right)}
\end{matrix}\right]
\end{align*}

Again, we can inspect these coordinate functions of $\boldsymbol \theta$ to look for a relationship between parameter values and uninterpretability for the toy RBM.
